### 2019-07-17 manuscript/abstract.tex
Small edits to space etc. in abstract and intro. Returned one sentence in each of two discussion paragraphs (the point of Ks timing is that the event is not within the family, which is critical for the section and inference evaluation.)

If particular traits consistently affect rates of speciation and extinction, broad macroevolutionary patterns can be understood as consequences of selection at high levels of the biological hierarchy.
Identifying traits associated with diversification rate differences is complicated by the wide variety of characters under consideration and the statistical challenges of testing for associations from comparative phylogenetic data.
Ploidy (diploid \vs polyploid states) and breeding system (self-incompatible \vs self-compatible states) have been repeatedly suggested as possible drivers of differential diversification.
We investigate the connections of these traits, including their interaction, to speciation and extinction rates in Solanaceae.
We show that the effect of ploidy on diversification can be largely explained by its correlation with breeding system and that additional unknown factors, alongside breeding system, determine diversification rates.
These results are largely robust to allowing for diploidization.
Finally, we find that the evolutionary pathway to polyploidy in Solanaceae is most often via the direct breakdown of self-incompatibility by whole genome duplication, rather than indirectly via breakdown followed by polyploidization.

>>>>

If particular traits consistently affect rates of speciation and extinction, broad macroevolutionary patterns can be understood as consequences of selection at high levels of the biological hierarchy.
Identifying traits associated with diversification rate differences is complicated by the wide variety of characters under consideration and the statistical challenges of testing for associations from comparative phylogenetic data.
Ploidy (diploid \vs polyploid states) and breeding system (self-incompatible \vs self-compatible states) have been repeatedly suggested as possible drivers of differential diversification.
We investigate the connections of these traits, including their interaction, to speciation and extinction rates in Solanaceae.
We show that the effect of ploidy on diversification can be largely explained by its correlation with breeding system and that additional unknown factors, alongside breeding system, determine diversification rates.
These results are largely robust to allowing for diploidization.
Finally, we find that the most common evolutionary pathway to polyploidy in Solanaceae occurs via direct breakdown of self-incompatibility by whole genome duplication, rather than indirectly via breakdown followed by polyploidization.

### 2019-07-17 manuscript/discussion.tex
Small edits to space etc. in abstract and intro. Returned one sentence in each of two discussion paragraphs (the point of Ks timing is that the event is not within the family, which is critical for the section and inference evaluation.)

% E: should this results summary be much shorter?
% B: I think it's fine, but I switched it to present tense (not strictly, but almost) for both reasons of style (aesthetic preference, perhaps), to separate previous work from current work, and to avoid awkward implications about the process not being ongoing. Moreover, the use of "show" or "demonstrate" is very loose in the context of our weak inference resulting from--what we all hopefully agree--are potentially flawed models.
Previous analyses of the effects of ploidy on diversification found that diploids are associated with greater net diversification rates than polyploids across many angiosperm clades \citep{mayrose_2011, mayrose_2015}. 
We obtain a similar outcome when examining ploidy alone in Solanaceae (\cref{figure:netdivall}A), but a consistent effect of ploidy on diversification is not clear when we incorporate a hidden factor linked to diversification (\cref{figure:netdivall}B).
%
Previous analyses of breeding system in this family found that SI may cause higher diversification rates, compared with SC (\citealt{goldberg_2010}; \cref{figure:netdivall}C).
Our analyses that include a hidden trait recover the same pattern, with one important difference.
On the background of one hidden state, we recover a net diversification rate for SC species that is positive and greater than the diversification rate of SI species on the background of the other hidden state (\cref{figure:netdivall}D).%B: I rewrote this to separate the point of the side clause. 
Therefore, SC may not be a `dead end' (cf.\ \citealt{igic_2013}) when coupled with some unknown trait combinations (or processes not modeled by these two traits). 
%Separately, I should note here that nothing about this analysis seems to me to strongly imply that SC is not a dead end in combination with a trait (as we discuss immediately below!). Much of the explanatory power derived for the A/B model may come from (for instance) accommodating tree scaling problems or rate heterogeneity in molecular evolution. It just fills up all of the mystery space without pointing to a cause. Side notes: (1) The I/C + A/B model ancestral states basically have the root of the tree as being in one hidden state and then tips in another. (2) Is there any place where we can see how much of the tree is in hidden state A or state B under some model?
%
Our analyses also reveal that models of joint evolution of ploidy and breeding system are statistically preferred, and hint at how various trait combinations may be linked with diversification.
We find that the highest net diversification rate is associated with SI diploids, while SC diploids have a lower diversification rate that overlaps with the net diversification of SC polyploids (\cref{figure:netdivall}E).
Thus, breeding system appears to provide a relatively better explanation of diversification rate differences, with ploidy providing a secondary effect within SC species.

@@

Polyploidization is known to be common in plants, but the relative frequency of the reverse process---diploidization---remains unclear, and it is under active investigation \citep{dodsworth_2015,mandakova_2018}.
%
Ignoring diploidization, if it is common, could cause underestimation and increase of uncertainty in  the polyploidization rate. 
Therefore, we included a slate of models with a diploidization parameter, and show that our main conclusions are robust to this process (\cref{suppfigure:alldip}).
These models also suggest modest statistical support for diploidization (\cref{supptable:testdiploidization}), although our estimates of its rate were highly uncertain. 
Furthermore, additional lines of evidence for classifying species as diploid or polyploid (beyond the genus-level chromosome multiplicity that we primarily relied on) are needed for more reliable conclusions.
%In the cases where ploidy level was assigned by us based chromosome multiplicity at the genus level, estimating diploidization might be a potential issue due to ploidy misclassification.  
%R: By adding diploidization to models of polyploidy linked to diversification it is possible to recover this complicated scenario and to reconcile genomic evidence with comparative phylogenetic models. 
%B: yielding to %R here. I still think there is overwhelming support for this statement, and few of the cited papers show convincingly that, for example, diploidization happens within five orders of magnitude compared with polyploidization, rendering "relatively rare" obsolete or invalid. (P.S. Rare not the same as unimportant!)
%B: was: Diploidization is widely considered to be relatively rare, compared with polyploidization \citep{husband_2013}, but flowering plant lineages are thought to have experienced at least one round of polyploidization in their evolutionary history \citep{soltis_2015}.

% E: Boris, could you please revisit the following two paragraphs?  TODO
% Two questions:
%    Did my reorganization of this section mess anything up?
%    Can this text be shorter?  It's seems disproportionately long, and I'm hesitant to revise it heavily myself because the logical connections aren't entirely clear to me.

Other lines of evidence about the prevalence of diploidization within Solanaceae or its ancestors are mixed or even conflicting.
% Part I: evidence for WGD in/near family
On the one hand, polyploidy may have occurred prior to the origin of Solanaceae, rendering all extant `diploids' secondarily derived.
\Citet{ku2000} and \citet{blanc2004} posited that the lineage leading to tomato, \emph{Solanum lycopersicum}, may have experienced one or more whole genome duplications.
A subsequent analysis of synteny between grape and \emph{Solanum} genomes, as well as the distribution between inferred paralogs within genomes of \emph{Solanum} (tomato and potato) each suggested that this lineage experienced a likely round of ancient genome duplication or triplication ( $71 \pm 19$ My, \citet{tomato2012})  likely pre-dating the origin of the family ($49 \pm 3$ My; \citealt{sarkinen_2013}).

%The age of the peak of paralog Ks distances is approximately $71 \pm 19$ My \citep{tomato2012}. 
% R- not entering into the weeds of Ks plots
%f this is the case, then all of the genomes may have been subsequently diploidized, yielding the widespread and frequent chromosome numbers in this and closely related families, n=11--12, all presently considered to be diploid \citep{robertson_2011}.

% Part II: evidence against WGD in/near family
On the other hand, there is little evidence for the occurrence of diploidization after the origin of Solanaceae itself.
% R- This is redundant now we said it above
%First, the Ks-inferred duplication \citep{tomato2012} likely pre-dates the origin of the family (49 My, HPD 46--53 My; \citealt{sarkinen_2013}). 
%Thus, the species in our analyses could be ancestrally polyploid, but would have diploidized before the root of the family.
Studies comparing map-based genome synteny within the family find no evidence for recent diploidization (extensively discussed in \citet{wu_2010a}).
Instead, simple genome re-arrangements appear sufficient to explain chromosomal evolution between a number of species. % including all of those in the relatively cytogenetically conserved x=12 group, which includes tomato, potato, eggplant, pepper, and tobacco.
Specifically, chromosome numbers and genome comparisons within the family (esp.\ the x=12 clade containing \textit{Solanum} and \textit{Nicotiana}) reveal strong conservation.

% R- removed the  inversion discussion I think the point was made above about the map base genome is sufficient. We don't need to describe more specifics

%\citet{wu_2010a} review the evidence from map-based genome comparisons and find that tomato and potato differ by as few as six inversions, tomato and eggplant by 24 inversions and five translocations, tomato and pepper by 19 inversions and six translocations, and tomato and tobacco by ca.\ 10 inversions and 11 translocations.
%Recovery of such simple relationships would require outstanding convergent loss of duplicated segments.
%
% Part IIb: evidence from SI [might need more explaining?]
Furthermore, whole genome duplication in a eudicot lineage ancestral to Solanaceae would clash with the evidence that the homologous mechanism of SI, which has been present continually in many families \citep{igic_2006}, breaks down nearly invariably in natural and induced tetraploids \citep{stone_2002, mcclure_2009}.
Most problematically in this context, it is unclear how to explain the maintenance of trans-generic polymorphism at the orthologous S-loci in Solanaceae and other families if SI was previously broken down by polyploidization.
%
Regardless of whether genome polyploidization, followed by widespread diploidization, is a dominant mechanism of genome evolution in Solanaceae, it is clear that more work is needed for a complete understanding of the joint evolution of ploidy and breeding systems.

% Flowering plant lineages are thought to have experienced at least one round of polyploidization in their evolutionary history \citep{soltis_2015}. 
% Following polyploidization, it is possible for genome reorganization, downsizing, and loss to occur \citep{dodsworth_2015, zenil_2016, mandakova_2018}.
% As a consequence, nearly all extant species classified as ``diploid" in our analyses are possibly secondary diploids, having undergone both polyploidization and subsequent diploidization.

% no discussion of magnitude of \delta vs \rho!

% problems: this system has a big set of advantages - carefully curated (co-)occurrences of two traits, genetic basis, etc. and everything is still contradictory and opaque.

@@

% E: The text previously here was great, but it read to me more like an abstract or methods/results.  I tried to make it more like a big-picture summary.  What other general points do we want to hit here?

>>>>

% E: should this results summary be much shorter?
Previous analyses of the effects of ploidy on diversification found that diploids are associated with greater net diversification rates than polyploids across many angiosperm clades \citep{mayrose_2011, mayrose_2015}. 
We obtain a similar outcome when examining ploidy alone in Solanaceae (\cref{figure:netdivall}A), but a consistent effect of ploidy on diversification is not clear when we incorporate a hidden factor linked to diversification (\cref{figure:netdivall}B).
%
Previous analyses of breeding system in this family found that SI may cause higher diversification rates, compared with SC (\citealt{goldberg_2010}; \cref{figure:netdivall}C).
Our analyses that include a hidden trait recover the same pattern, with one important difference.
On the background of one hidden state, we recover a net diversification rate for SC species that is positive and greater than the diversification rate of SI species on the background of the other hidden state (\cref{figure:netdivall}D).%B: I rewrote this to separate the point of the side clause. 
Therefore, SC may not be a `dead end' (cf.\ \citealt{igic_2013}) when coupled with some unknown trait combinations (or processes not modeled by these two traits). 
%Separately, I should note here that nothing about this analysis seems to me to strongly imply that SC is not a dead end in combination with a trait (as we discuss immediately below!). Much of the explanatory power derived for the A/B model may come from (for instance) accommodating tree scaling problems or rate heterogeneity in molecular evolution. It just fills up all of the mystery space without pointing to a cause. Side notes: (1) The I/C + A/B model ancestral states basically have the root of the tree as being in one hidden state and then tips in another. (2) Is there any place where we can see how much of the tree is in hidden state A or state B under some model?
%
Our analyses also reveal that models of joint evolution of ploidy and breeding system are statistically preferred, and hint at how various trait combinations may be linked with diversification.
We find that the highest net diversification rate is associated with SI diploids, while SC diploids have a lower diversification rate that overlaps with the net diversification of SC polyploids (\cref{figure:netdivall}E).
Thus, breeding system appears to provide a relatively better explanation of diversification rate differences, with ploidy providing a secondary effect within SC species.

@@

Polyploidization is known to be common in plants, but the relative frequency of the reverse process---diploidization---remains unclear, and it is under active investigation \citep{dodsworth_2015,mandakova_2018}.
%
Ignoring diploidization, if it is common, could cause underestimation and increase of uncertainty in  the polyploidization rate. 
Therefore, we included a slate of models with a diploidization parameter, and show that our main conclusions are robust to this process (\cref{suppfigure:alldip}).
These models also suggest modest statistical support for diploidization (\cref{supptable:testdiploidization}), although our estimates of its rate were highly uncertain. 
Furthermore, additional lines of evidence for classifying species as diploid or polyploid (beyond the genus-level chromosome multiplicity that we primarily relied on) are needed for more reliable conclusions.

Other lines of evidence about the prevalence of diploidization within Solanaceae or its ancestors are mixed or even conflicting.
% Part I: evidence for WGD in/near family
On the one hand, polyploidy may have occurred prior to the origin of Solanaceae, rendering all extant `diploids' secondarily derived.
\Citet{ku2000} and \citet{blanc2004} posited that the lineage leading to tomato, \emph{Solanum lycopersicum}, may have experienced one or more whole genome duplications.
A subsequent analysis of synteny between grape and \emph{Solanum} genomes, as well as the distribution between inferred paralogs within genomes of \emph{Solanum} (tomato and potato) each suggested that this lineage experienced a likely round of ancient genome duplication or triplication ( $71 \pm 19$ My, \citet{tomato2012})  likely pre-dating the origin of the family ($49 \pm 3$ My; \citealt{sarkinen_2013}). 
The age of the peak of paralog Ks distances is approximately $71 \pm 19$ My \citep{tomato2012}. 
%B: We cannot selectively delete the weak evidentiary basis for ku2000, blanc2004, and tomato2012. It's the only mention of their inference date and method, and it is a single sentence. %R- not entering into the weeds of Ks plots

% Part II: evidence against WGD in/near family
On the other hand, there is little evidence for the occurrence of diploidization after the origin of Solanaceae itself.
First, the Ks-inferred duplication \citep{tomato2012} pre-dates the origin of the family (49 My, HPD 46--53 My; \citealt{sarkinen_2013}). 
Studies comparing map-based genome synteny within the family find no evidence for recent diploidization \citep{wu_2010a}.
Instead, simple genome re-arrangements appear sufficient to explain chromosomal evolution within species in the cytogenetically conserved `x=12' group, which includes tomato, potato, eggplant, pepper, and tobacco.
Recovery of comparatively few rearrangements would require outstanding convergent loss of duplicated segments.
Furthermore, whole genome duplication in a eudicot lineage ancestral to Solanaceae would clash with the evidence that the homologous mechanism of SI, which has been present continually in many families \citep{igic_2006}, breaks down nearly invariably in natural and induced tetraploids \citep{stone_2002, mcclure_2009}.
Most problematically in this context, it is unclear how to explain the maintenance of trans-generic polymorphism at the orthologous S-loci in Solanaceae and other families if SI was previously broken down by polyploidization.
%
Regardless of whether genome polyploidization, followed by widespread diploidization, is a dominant mechanism of genome evolution in Solanaceae, it is clear that more work is needed for a complete understanding of the joint evolution of ploidy and breeding systems.

@@

### 2019-07-17 manuscript/introduction.tex
Small edits to space etc. in abstract and intro. Returned one sentence in each of two discussion paragraphs (the point of Ks timing is that the event is not within the family, which is critical for the section and inference evaluation.)

Species accumulate across the tree of life at different rates. 
One possible explanation for this phenomenon is that various traits differentially affect rates of diversification. 
Dramatic increases in phylogenetic and phenotypic data along with methodological advances have greatly accelerated the search for traits that influence diversification. 
Nevertheless, identifying focal traits associated with rates of speciation and extinction remains a challenge \citep[\eg][]{maddison_2015, rabosky_2015, beaulieu_2016, rabosky_2017}. 
One difficulty is that speciation and extinction likely do not depend on a single character, so the biological and environmental contexts in which traits occur can lead to complex interactions that affect lineage diversification \citep{beaulieu_2016, caetano_2018, herrera_2018}.
Consequently, examining the association of only one character with diversification patterns can be misleading. 
Here, we embrace this challenge by jointly investigating two characters thought to influence speciation and extinction rates---ploidy level and breeding system---while allowing for their interactions and other confounding factors to change diversification. We also test whether adding one more trait and increasing model complexity is worthwhile.
% E: We need a "Here, we..." here.  Otherwise, we've just told the reader "not one trait!" and then launch into one trait.
% Could also state that we newly test whether adding one more trait is worthwhile.
%B: I think you are correct, but it introduces a new concern: we now say N=1 is too few, so not why try N>2 characters? Why is N=2 the Goldilocks number? I didn't change anything but it's clear fodder for Discussion.
% R. added Emma's suggetion

@@

Breeding system shifts---changes in the collection of physiological and morphological traits that determine the likelihood that any two gametes unite---are also remarkably common and crucially affect the distribution and amount of genetic variation in populations \citep{barrett2013}. %B: note: we use "remarkably" to describe the frequency of both characters; hence "also."
In particular, a variety of genetic self-incompatibility (SI) systems cause plants to reject their own pollen, and loss of such mechanisms, yielding self-compatibility (SC), is one of the most replicated transitions in flowering plant evolution \citep{stebbins1974, igic_2008}.
Previous analyses reported higher rates of diversification for SI than for SC lineages in Solanaceae \citep{goldberg_2010}. 
Similarly, heterostylous SI lineages in Primulaceae seem to diversify faster \citep{devos2014}, as do outcrossing lineages in Onagraceae \citep{freyman_2019}.
Although these findings suggest a consistent macroevolutionary role of breeding system, it is unlikely to be the sole character determining lineage diversification.
We investigate the relationship of breeding system to speciation and extinction rates in the context of ploidy and other unobserved factors.

@@

In the present study, we employ an extended framework of state speciation and extinction models, which simultaneously model transitions between the discrete states of a trait and different rates of speciation and extinction associated with each of those states \citep[`SSE' models;][]{maddison_2007, fitzjohn_2012}.
% All the models proposed for this study are coded as graphical models and Bayesian inference was performed over the proposed models using RevBayes software \citep{hoehna_2016}.  % E: The models aren't drawn graphically in any of the figures (not in the revbayes sense, at least).  And we don't need methods in the intro.
% R- The graphical model coding is not about the figures but about how you code the nodes in the code. But I agree that it might be confusing especiallly for most botanists.
%Although potentially very powerful, this approach can be prone to reporting a false association between a trait and lineage diversification when the data contain additional, unmodeled diversification rate heterogeneity \citep{maddison_2015, rabosky_2015}.
We start by fitting binary state speciation and extinction models to each ploidy and breeding system independently \citep{maddison_2007}.
We follow by fitting models that incorporate hidden states, which reduce the chance that an effect of one trait on diversification would be mistaken for the effect of an unobserved factor \citep{beaulieu_2016}.
We compare the proposed models against their character independent counterparts \citep{beaulieu_2016} to investigate whether ploidy or breeding system are actually informative of the diversification process.
Next, we model ploidy and breeding system jointly to assess their combined influence on diversification, with or without an additional hidden character.
Using the ploidy and breeding system model without hidden states, we quantify the relative contributions of the two pathways from SI diploids to SC polyploids.
We also aggregate  states within these joint models of ploidy and breeding system in order to test whether increasing complexity from one trait to two traits significantly improves our understanding of the diversification process.
%B: I altered this line, but it is still a bit redundant given previous sentence.R-fixed
%Our analyses evaluate combinations of macroevolutionary models to parse out the associations of ploidy and breeding system on speciation and extinction rates in Solanaceae, while simultaneously accounting for other processes that influence diversification.
%First, we present models in which the effects of ploidy and breeding system are considered separately and compare the inferences to previously published results. 
%Second, we add a hidden character to each model and investigate if state-dependent diversification in focal character retains its explanatory power, or whether the added hidden character obscures it.
Furthermore, we extend all the models involving ploidy to investigate the potential effects of including diploidization.
% And finally, we explore models of diversification using asymmetric rates between and within hidden states to investigate the sensitivity of the statistical inference to model assumptions.
Our results highlight the importance of considering non-additive effects of traits on net diversification rates  under the presence of unobserved factors, in order to detect strong biologically-driven processes dictating the diversification patterns.

>>>>

Species accumulate across the tree of life at different rates. 
One possible explanation for this phenomenon is that various traits differentially affect rates of diversification. 
Dramatic increases in phylogenetic and phenotypic data, along with methodological advances, have greatly accelerated the search for traits that influence diversification.
Nevertheless, identifying focal traits associated with rates of speciation and extinction remains a challenge \citep[\eg][]{maddison_2015, rabosky_2015, beaulieu_2016, rabosky_2017}. 
One difficulty is that speciation and extinction likely do not depend on a single character, so the biological and environmental contexts in which traits occur can lead to complex interactions that affect lineage diversification \citep{beaulieu_2016, caetano_2018, herrera_2018}.
Consequently, examining the association of only one character with diversification patterns can be misleading. 
Here, we embrace this challenge by jointly investigating two characters thought to influence speciation and extinction rates---ploidy level and breeding system---while allowing for their interactions, and other confounding factors, to change diversification. 
We also test whether adding one more trait and increasing model complexity is worthwhile.

@@

Breeding system shifts---changes in the collection of physiological and morphological traits that determine the likelihood that any two gametes unite---are also remarkably common and crucially affect the distribution and amount of genetic variation in populations \citep{barrett2013}. 
In particular, a variety of genetic self-incompatibility (SI) systems cause plants to reject their own pollen, and loss of such mechanisms, yielding self-compatibility (SC), is one of the most replicated transitions in flowering plant evolution \citep{stebbins1974, igic_2008}.
Previous analyses reported higher rates of diversification for SI than for SC lineages in Solanaceae \citep{goldberg_2010}. 
Similarly, heterostylous SI lineages in Primulaceae seem to diversify faster \citep{devos2014}, as do outcrossing lineages in Onagraceae \citep{freyman_2019}.
Although these findings suggest a consistent macroevolutionary role of breeding system, it is unlikely to be the sole character determining lineage diversification.
We investigate the relationship of breeding system to speciation and extinction rates in the context of ploidy and other unobserved factors.

@@

In the present study, we employ an extended framework of state speciation and extinction models, which simultaneously model transitions between the discrete states of a trait and different rates of speciation and extinction associated with each of those states \citep[`SSE' models;][]{maddison_2007, fitzjohn_2012}.
We start by fitting binary state speciation and extinction models to each ploidy and breeding system independently \citep{maddison_2007}.
We follow by fitting models that incorporate hidden states, which reduce the chance that an effect of one trait on diversification would be mistaken for the effect of an unobserved factor \citep{beaulieu_2016}.
We compare the proposed models against their character independent counterparts \citep{beaulieu_2016} to investigate whether ploidy or breeding system are actually informative of the diversification process.
Next, we model ploidy and breeding system jointly to assess their combined influence on diversification, with or without an additional hidden character.
Using the ploidy and breeding system model without hidden states, we quantify the relative contributions of the two pathways from SI diploids to SC polyploids.
We also aggregate states within these joint models of ploidy and breeding system in order to test whether increasing complexity from one trait to two traits significantly improves our understanding of the diversification process.
Furthermore, we extend all the models involving ploidy to investigate the potential effects of including diploidization.
% And finally, we explore models of diversification using asymmetric rates between and within hidden states to investigate the sensitivity of the statistical inference to model assumptions.
Our results highlight the importance of considering non-additive effects of traits on net diversification rates  under the presence of unobserved factors, in order to detect strong biologically-driven processes dictating the diversification patterns.
